---
title: Project Proposal for Modelling Democracy
authors: Moksha Sharma, Divij Sinha
format: html
---

```{r}
library(tidyverse)
```

```{r}
shrug_triv_elect <- read_csv("data/shrug-triv-elect-csv/trivedi_elections_clean.csv")
shrug_triv_cand <- read_csv("data/shrug-triv-cand-csv/trivedi_candidates_clean.csv")
shrug_adr_elect <- read_csv("data/shrug-adr-elections-ac-csv/affidavits_ac.csv")
shrug_adr_cand <- read_csv("data/shrug-adr-elections-cand-csv/affidavits_clean.csv")
```

`shrug_triv_elect` -> Election level with winner info. Each row is one election (one state-year-seat)
`shrug_triv_cand` -> Candidate level with candidate info. Each row is one election candidate (one state-year-seat-candidate)
`shrug_adr_elect` -> Election level with avg affidavit info. Each row is one election (one state-year-seat)
`shrug_adr_cand` -> Candidate level with candidate affidavit info. Each row is one candidate (one state-year-seat-candidate)

## Total elections with NOTA

Total number of elections

```{r}
print(nrow(shrug_triv_elect)) ## Total Elections in election data
print(nrow(shrug_triv_cand %>% select(sh_election_id) %>% unique())) ## Total Elections in candidate data
```

Total number of non bye-poll elections 

```{r}
print(nrow(shrug_triv_elect %>% filter(bye_election == 0))) ## Total Elections in election data
print(nrow(shrug_triv_cand %>% filter(bye_election == 0) %>% select(sh_election_id) %>% unique())) ## Total Elections in candidate data
```

All non bye-poll elections with nota flag

```{r}
election_is_nota <- shrug_triv_cand %>%
    filter(bye_election == 0) %>%
    group_by(sh_election_id, year) %>%
    summarise(is_nota = any(tolower(cand_name) == "nota" | tolower(cand_name) == "none of the above"), .groups = "drop")

head(election_is_nota)
```

Summary table

```{r}
election_is_nota %>%
    group_by(year) %>%
    summarise(
        perc_nota_elections = mean(is_nota),
        count_nota_elections = sum(is_nota),
        count_total_elections = n(),
    ) %>%
    print(n = 50)
```

## Merging with affidavit 

```{r}
print(nrow(shrug_adr_elect)) ## Total Elections in affidavit data
print(nrow(shrug_adr_cand %>% select(sh_election_id) %>% unique())) ## Total Elections in affidavit election data
```

Total number of non bye-poll elections 

```{r}
print(nrow(shrug_adr_elect %>% filter(bye_election == 0))) ## Total Elections in affidavit data
print(nrow(shrug_adr_cand %>% filter(bye_election == 0) %>% select(sh_election_id) %>% unique())) ## Total Elections in affidavit election data
```

### Election level

```{r}
adr_elect <- inner_join(shrug_adr_elect, shrug_triv_elect) %>% inner_join(election_is_nota)

adr_elect %>%
    group_by(year) %>%
    summarise(
        perc_nota_elections = mean(is_nota),
        count_nota_elections = sum(is_nota),
        count_total_elections = n(),
    ) %>%
    print(n = 50)
```

### Candidate level

```{r}
adr_cand <- inner_join(shrug_adr_cand, shrug_triv_cand, by = join_by("sh_cand_id", "sh_election_id"), suffix = c("_adr", ""), ) %>% inner_join(election_is_nota)

nrow(adr_cand)
```

```{r}
adr_cand %>%
    group_by(year) %>%
    summarise(
        mean_crimes = mean(num_crim),
        median_crimes = median(num_crim),
        max_crimes = max(num_crim),
        perc_crime_major = mean(crime_major, na.rm = T),
        perc_crime_major_adr = mean(crime_major_adr, na.rm = T)
    )
```