---
title: Project Proposal for Modelling Democracy
authors: Moksha Sharma, Divij Sinha
format: html
bibliography: references.bib
execute:
  cache: true
---

# Introduction

Voting is costly: eligible voters need to collect information on candidates, formulate their preferences, and find the time to express these preferences at the ballot. Given these costs, political scientists have offered many explanations for why citizens make the effort to vote despite their vanishingly small chances of casting the pivotal vote. Existing explanations focus on <civic duty, expressiveness, social pressure?>. However, these explanations fail to explain why people would take the time to familiarise themselves with the candidates, stand in line, and cast their vote, …

Many countries offer a similar option to vote for no candidate listed on the ballot, but they impose different requirements for what to do with the outcomes of an election in which NOTA receives a substantial share of the vote. In Colombia and Bangladesh, for instance, elections need to be repeated if NOTA receives the absolute majority of the valid ballots. On the other hand, in India and Bulgaria, the votes are counted but not factored into the outcome of the elections. In other words, while NOTA allows voters in these countries to formally express their discontentment with the candidates, this discontentment is toothless: there are no formal requirements for repeating elections or banning candidates from running again. 

NOTA is a distinctive modification to plurality. Unlike adding a candidate, NOTA accumulates votes but can never enter a winning coalition — structurally, it behaves as a dummy player. Even if NOTA receives the most votes, the leading candidate wins. (A few Municipal Elections have adopted re-election provisions when NOTA wins a majority, but no such rule applies to PC or AC races.) This creates an unusual dynamic: NOTA gives voters an expressive channel with no direct mechanical consequence for the outcome.

The question we ask is whether this non-competitive option has *indirect* consequences on the competitive landscape of elections, or changes how voters weigh candidate characteristics against the closeness of the race.

# Research Question

This project draws on the rational voter framework from @barnett2020you, as discussed in the course. Barnett establishes that voting is individually rational when two conditions hold: a *stakes condition* (the social benefit of your preferred candidate exceeds twice the cost of voting) and a *chances condition* (the probability of being decisive is at least $1/N$). The chances condition is more likely to bind in competitive races with narrow margins.

We break our question into two parts that map onto these conditions.

## Part 1: Does NOTA Change the Competitive Landscape?

First, we ask whether NOTA has shifted the competitiveness of Indian elections — specifically, whether margins of victory have narrowed or widened, and whether voter turnout has changed. If NOTA draws votes away from leading candidates, it could mechanically tighten margins, pushing more constituencies toward the regime where the chances condition binds and individual votes are more likely to be decisive. Alternatively, if NOTA primarily absorbs would-be abstainers, it would increase turnout without affecting margins.

Distinguishing between these scenarios tells us something about what NOTA is doing to the electorate. In the first case, NOTA could be functioning almost like a spoiler (since NOTA cannot win but its presence might change the winner set, by siphoning votes away from the "natural" winner). In the second case, NOTA is simply converting abstention into a recorded preference, revealing dissatisfaction that was previously invisible.

## Part 2: Criminality, Voter Preference, and Strategic Tolerance

Second, we investigate how voters weigh candidate criminality against electoral competitiveness. Indian candidates are required to file affidavits disclosing pending criminal cases (along with net worth, and education), creating rich data on candidate quality.

We start by estimating the effect of additional criminal cases against a candidate on their vote share, and on the NOTA vote share, measuring voter sensitivity to candidate quality, and how likely a voter is to move towards NOTA given some increase in candidate disquality. Next, we ask whether these sensitivities depend on the competitiveness of the race. In a tight election where a voter's expected influence is higher, which under the rational voter framework could cut either way: voters might "hold their nose" and support a flawed but preferred candidate because the stakes of the outcome are too high, or they might be *more* willing to defect to NOTA precisely because their vote is more visible in a close race.
 
# Analysis

```{r setup}
#| warning: false
#| message: false
#| output: false
library(tidyverse)
library(fixest)
shrug_triv_elect <- read_csv("data/shrug-triv-elect-csv/trivedi_elections_clean.csv")
shrug_triv_cand <- read_csv("data/shrug-triv-cand-csv/trivedi_candidates_clean.csv")
shrug_adr_elect <- read_csv("data/shrug-adr-elections-ac-csv/affidavits_ac.csv")
shrug_adr_cand <- read_csv("data/shrug-adr-elections-cand-csv/affidavits_clean.csv")
```

`shrug_triv_elect` -> Election level with winner info. Each row is one election (one state-year-seat)  
`shrug_triv_cand` -> Candidate level with candidate info. Each row is one election candidate (one state-year-seat-candidate)  
`shrug_adr_elect` -> Election level with avg affidavit info. Each row is one election (one state-year-seat)  
`shrug_adr_cand` -> Candidate level with candidate affidavit info. Each row is one candidate (one state-year-seat-candidate)

## Basic Data Landscape

### Elections

We have access to state election level data for election results and candidate level affidavits. 

We first start by looking at the election level data. These include elections from 1979 to 2022.

```{r}
#| results: hold
print(nrow(shrug_triv_elect)) ## Total Elections in election data
print(nrow(shrug_triv_cand %>% select(sh_election_id) %>% unique())) ## Total Elections in candidate data
```

```{r}
#| results: hold
print(nrow(shrug_triv_elect %>% filter(bye_election == 0))) ## Total Elections in election data
print(nrow(shrug_triv_cand %>% filter(bye_election == 0) %>% select(sh_election_id) %>% unique())) ## Total Elections in candidate data
```

Therefore, we have a total of ~43,000 elections, which upon filtering to remove bye-poll elections leaves us with ~41,600 elections. 

We remove bye-polls because they usually have different, smaller sets of candidates and compaigns, and have no compulsion to have "None of the Above" as an option.


```{r}
election_is_nota <- shrug_triv_cand %>%
    filter(bye_election == 0) %>%
    group_by(sh_election_id, year) %>%
    summarise(is_nota = any(tolower(cand_name) == "nota" | tolower(cand_name) == "none of the above"), .groups = "drop")

election_is_nota %>%
    group_by(year) %>%
    summarise(
        perc_nota_elections = mean(is_nota),
        count_nota_elections = sum(is_nota),
        count_total_elections = n(),
    ) %>%
    print(n = 50)
```

Within these, as we can see, the elections with NOTA as an option start happening in 2013, and barring one bad data point, exist in every election since.

### Affidavits

<!-- 
```{r}
#| results: hold
print(nrow(shrug_adr_elect)) ## Total Elections in affidavit data
print(nrow(shrug_adr_cand %>% select(sh_election_id) %>% unique())) ## Total Elections in affidavit election data
```

```{r}
#| results: hold
print(nrow(shrug_adr_elect %>% filter(bye_election == 0))) ## Total Elections in affidavit data
print(nrow(shrug_adr_cand %>% filter(bye_election == 0) %>% select(sh_election_id) %>% unique())) ## Total Elections in affidavit election data
``` 
-->

Now, looking at the affidavit level data, we have information on affidavits from 2004 to 2017. While this limits the election data we are able to use, as we can see, it still gives us about 11,000 elections we can still analyse.

**Merging election level affidavit data**

```{r}
#| results: hold
#| message: false
adr_elect <- inner_join(shrug_adr_elect, shrug_triv_elect) %>% inner_join(election_is_nota)

adr_elect %>%
    group_by(year) %>%
    summarise(
        perc_nota_elections = mean(is_nota),
        count_nota_elections = sum(is_nota),
        count_total_elections = n(),
    ) %>%
    print(n = 50)
```

```{r}
adr_elect %>% group_by(is_nota) %>% summarise(n = n())
```

**Merging Candidate level affidavit data**

```{r}
#| message: false
adr_cand <- left_join(shrug_triv_cand, shrug_adr_cand, by = join_by("sh_cand_id", "sh_election_id"), suffix = c("", "_adr"), ) %>%
    inner_join(election_is_nota)

adr_cand %>%
    filter(!is.na(adr_cand_id)) %>%
    group_by(year) %>%
    summarise(
        mean_crimes = mean(num_crim, na.rm = T),
        median_crimes = median(num_crim, na.rm = T),
        max_crimes = max(num_crim, na.rm = T),
        perc_crime_major = mean(crime_major, na.rm = T),
        perc_crime_major_adr = mean(crime_major_adr, na.rm = T)
    ) %>%
    round(2)
```

## Estimations 

**Average NOTA share by state-year**

```{r}
#| message: false
adr_cand <- adr_cand %>%
    mutate(nota_cand = ifelse(tolower(cand_name) %in% c("nota", "none of the above"), 1, 0))

adr_cand %>%
    filter(nota_cand == 1) %>%
    group_by(eci_state_name, year) %>%
    summarise(
        median_vsp = median(vote_share_percentage),
        mean_vsp = mean(vote_share_percentage),
    ) %>%
    print(n=40)
```

At the state level, we can see that average percentage of votes polled as NOTA hoever around 1%, with interesting outliers in the case of the states of Bihar and Chhattisgarh - both poorer states with a (at least perceived) higher degree of corruption and maladministration.

```{r}
#| message: false
#| warning: false
election_nota_votes <- adr_cand %>%
    filter(nota_cand == 1) %>%
    mutate(
        nota_votes = votes,
        nota_vote_share = votes / valid_votes
    ) %>%
    select(sh_election_id, nota_votes, nota_vote_share)

ggplot(data = election_nota_votes, aes(x = nota_vote_share * 100)) +
    geom_density() +
    geom_vline(xintercept = 1, color = "red") +
    geom_vline(xintercept = 2, color = "red") +
    annotate("text", x = 1.1, y = 1, label = "1%", size = 3) +
    annotate("text", x = 2.1, y = 1, label = "2%", size = 3) +
    labs(
        x = "NOTA vote share (%)",
        y = "Density",
        title = "Distribution of NOTA vote share across elections"
    ) +
    theme_minimal()
```

**Constituency-year with NOTA votes higher than winning margin**

```{r}
#| message: false

win_margin <- adr_cand %>%
    arrange(sh_election_id, desc(votes)) %>%
    group_by(sh_election_id) %>%
    mutate(
        win_margin = max(votes) - votes,
        win_margin_perc = 100 * (win_margin / valid_votes)
    ) %>%
    filter(win_margin != 0) %>%
    group_by(sh_election_id) %>%
    summarise(
        win_margin = min(win_margin),
        win_margin_perc = min(win_margin_perc)
    )

nota_margin <- win_margin %>%
    inner_join(election_nota_votes) %>%
    mutate(diff_votes = win_margin - nota_votes) %>%
    arrange(diff_votes)

nota_margin %>% filter(diff_votes <= 0)
```

Surprisingly, of the 7,544 elections that have NOTA as an option, 474 or over 6% of the elections have a margin of victory that is less than the NOTA votes polled!

This points to the fact that some of these elections could actually be quite sensitive to the NOTA vote.

The NOTA vote represents something unique, it is someone taking the effort to the voting booth, stand in line for potentially hours, and then in some sense reveal a preference in a manner that is not really "relevant" to the election result. It is a costly signal with little to no benefit.

Let's now look at turnout trends, and vote-margins, pre and post NOTA

```{r}
shrug_triv_elect %>%
    filter(year >= 2004) %>%
    inner_join(election_is_nota) %>%
    feols(fml = turnout_percentage ~ is_nota | year + pc01_state_id + tr_ac_id, data = .) %>%
    summary()
```

As we can see, NOTA seems to have a weak impact on turnout trends, looking at the 10 year window before and after the introduction of the change!

```{r}
shrug_triv_elect %>%
    filter(year >= 2004) %>%
    inner_join(election_is_nota) %>%
    inner_join(win_margin) %>%
    feols(fml = win_margin_perc ~ is_nota | year + pc01_state_id + tr_ac_id, data = .) %>%
    summary()
```

However, NOTA does not seem to impact the average win margin.

```{r}
shrug_triv_elect %>%
    filter(year >= 2004) %>%
    inner_join(election_is_nota) %>%
    feols(fml = n_cand ~ is_nota | year + pc01_state_id + tr_ac_id, data = .) %>%
    summary()
```

More importantly, it seems to reduce the average number of candidates in the election by 2!

The number of candidates is already quite high, with a median above 10, but since most of candidates are not competitive, this is showing us that since there is already a sink for protest votes, protest candidates, rarely, if at all run for election now that NOTA is an option.

NOTE: The rest is still WIP

```{r}
ggplot(data = nota_margin %>% filter(abs(diff_votes) < 500)) +
    geom_histogram(aes(x = diff_votes), binwidth = 5)
```


However, if elections are close, people seem to be voting for NOT nota so the election has more win margin(??)

<!-- 
**Distribution of NOTA vote share**

```{r}
#| message: false
ggplot(data = nota_margin, aes(x = nota_vote_share * 100)) +
    geom_density(fill = "steelblue", alpha = 0.4, color = "steelblue4") +
    geom_vline(xintercept = 1, linetype = "dashed", color = "firebrick", linewidth = 0.7) +
    geom_vline(xintercept = 2, linetype = "dotted", color = "firebrick", linewidth = 0.7) +
    annotate("text", x = 1.1, y = Inf, label = "1%", vjust = 2, hjust = 0, color = "firebrick", size = 3) +
    annotate("text", x = 2.1, y = Inf, label = "2%", vjust = 2, hjust = 0, color = "firebrick", size = 3) +
    labs(
        x = "NOTA vote share (%)",
        y = "Density",
        title = "Distribution of NOTA vote share across elections"
    ) +
    theme_minimal()
```

The distribution is heavily right-skewed: the vast majority of elections see NOTA below 2%, but a non-trivial tail extends beyond that. The 1% and 2% reference lines put the "spoiler zone" in context — the elections flagged above as margin < NOTA votes live in this right tail.
-->

<!-- **NOTA vote share vs. winning margin**

```{r}
#| message: false
nota_margin_plot <- nota_margin %>%
    mutate(
        nota_vote_share_perc = nota_vote_share * 100,
        spoiler = diff_votes <= 0
    )

ggplot(data = nota_margin_plot, aes(x = nota_vote_share_perc, y = win_margin_perc, color = spoiler)) +
    geom_point(alpha = 0.3, size = 0.8) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black", linewidth = 0.8) +
    scale_color_manual(
        values = c("FALSE" = "steelblue", "TRUE" = "firebrick"),
        labels = c("FALSE" = "Margin > NOTA", "TRUE" = "Margin ≤ NOTA"),
        name = NULL
    ) +
    coord_cartesian(xlim = c(0, 8), ylim = c(0, 20)) +
    labs(
        x = "NOTA vote share (%)",
        y = "Win margin (% of valid votes)",
        title = "NOTA vote share vs. winning margin",
        subtitle = "Elections below the diagonal line have margin ≤ NOTA votes"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
```

Elections in red fall below the diagonal — their margin of victory was smaller than the NOTA vote. These are the elections where NOTA was structurally decisive, even though it mechanically could not affect the outcome. The clustering of most elections well above the diagonal confirms that this is a tail phenomenon, but the red points establish that it is not negligible.
 -->

## Other tables 

Criminal affidavits per const


```{r}
election_crim <- adr_cand %>%
    filter(is_nota == 1) %>%
    group_by(sh_election_id) %>%
    summarise(
        num_criminals = sum(num_crim > 0, na.rm = TRUE),
        median_num_crim = median(num_crim, na.rm = TRUE),
        mean_num_crim = mean(num_crim, na.rm = TRUE)
    ) %>%
    inner_join(election_nota_votes)


cm_nota <- election_crim %>%
    group_by(median_num_crim) %>%
    summarise(
        median_nota_vote_share = median(nota_vote_share, na.rm = TRUE),
        mean_nota_vote_share = mean(nota_vote_share, na.rm = TRUE)
    )


ggplot(data = cm_nota) +
    geom_col(aes(x = median_num_crim, y = median_nota_vote_share))
```


```{r}
major_crim <- adr_cand %>%
    filter(is_nota == 1) %>%
    group_by(sh_election_id) %>%
    summarise(
        crime_major = sum(crime_major_adr, na.rm = TRUE),
    ) %>%
    inner_join(election_nota_votes)


cm_nota <- major_crim %>%
    group_by(crime_major) %>%
    summarise(
        median_nota_vote_share = median(nota_vote_share, na.rm = TRUE),
        mean_nota_vote_share = mean(nota_vote_share, na.rm = TRUE)
    )


ggplot(data = cm_nota) +
    geom_col(aes(x = crime_major, y = median_nota_vote_share))
```


# Data

The data for this project comes from the following sources:

- PC and AC election results sourced from Lok Dhaba, TCPD, Ashoka University (@agarwal2021tcpd)
- Candidate affidavit data sourced from [myneta.info](myneta.info) by the Association for Democratic Reforms (ADR)


# References

::: {#refs .references}
:::