---
title: Project Proposal for Modelling Democracy
authors: Moksha Sharma, Divij Sinha
format: html
bibliography: references.bib
execute:
  cache: true
---

# Introduction

India's Parliamentary Constituency (PC) and Assembly Constituency (AC) elections use plurality rule — each voter casts a single first-preference vote, and the candidate with the most votes wins. In 2013, India introduced "None of the Above" (NOTA) as a formal ballot option in all direct PC and AC elections.

NOTA is a distinctive modification to plurality. Unlike adding a candidate, NOTA accumulates votes but can never enter a winning coalition — structurally, it behaves as a dummy player. Even if NOTA receives the most votes, the leading candidate wins. (A few Municipal Elections have adopted re-election provisions when NOTA wins a majority, but no such rule applies to PC or AC races.) This creates an unusual dynamic: NOTA gives voters an expressive channel with no direct mechanical consequence for the outcome.

The question we ask is whether this non-competitive option has *indirect* consequences on the competitive landscape of elections, or changes how voters weigh candidate characteristics against the closeness of the race.

# Research Question

This project draws on the rational voter framework from @barnett2020you, as discussed in the course. Barnett establishes that voting is individually rational when two conditions hold: a *stakes condition* (the social benefit of your preferred candidate exceeds twice the cost of voting) and a *chances condition* (the probability of being decisive is at least $1/N$). The chances condition is more likely to bind in competitive races with narrow margins.

We break our question into two parts that map onto these conditions.

## Part 1: Does NOTA Change the Competitive Landscape?

First, we ask whether NOTA has shifted the competitiveness of Indian elections — specifically, whether margins of victory have narrowed or widened, and whether voter turnout has changed. If NOTA draws votes away from leading candidates, it could mechanically tighten margins, pushing more constituencies toward the regime where the chances condition binds and individual votes are more likely to be decisive. Alternatively, if NOTA primarily absorbs would-be abstainers, it would increase turnout without affecting margins.

Distinguishing between these scenarios tells us something about what NOTA is doing to the electorate. In the first case, NOTA could be functioning almost like a spoiler (since NOTA cannot win but its presence might change the winner set, by siphoning votes away from the "natural" winner). In the second case, NOTA is simply converting abstention into a recorded preference, revealing dissatisfaction that was previously invisible.

## Part 2: Criminality, Voter Preference, and Strategic Tolerance

Second, we investigate how voters weigh candidate criminality against electoral competitiveness. Indian candidates are required to file affidavits disclosing pending criminal cases (along with net worth, and education), creating rich data on candidate quality.

We start by estimating the effect of additional criminal cases against a candidate on their vote share, and on the NOTA vote share, measuring voter sensitivity to candidate quality, and how likely a voter is to move towards NOTA given some increase in candidate disquality. Next, we ask whether these sensitivities depend on the competitiveness of the race. In a tight election where a voter's expected influence is higher, which under the rational voter framework could cut either way: voters might "hold their nose" and support a flawed but preferred candidate because the stakes of the outcome are too high, or they might be *more* willing to defect to NOTA precisely because their vote is more visible in a close race.
 
# Analysis

```{r setup}
#| warning: false
#| message: false
#| output: false
library(tidyverse)
shrug_triv_elect <- read_csv("data/shrug-triv-elect-csv/trivedi_elections_clean.csv")
shrug_triv_cand <- read_csv("data/shrug-triv-cand-csv/trivedi_candidates_clean.csv")
shrug_adr_elect <- read_csv("data/shrug-adr-elections-ac-csv/affidavits_ac.csv")
shrug_adr_cand <- read_csv("data/shrug-adr-elections-cand-csv/affidavits_clean.csv")
```

`shrug_triv_elect` -> Election level with winner info. Each row is one election (one state-year-seat)  
`shrug_triv_cand` -> Candidate level with candidate info. Each row is one election candidate (one state-year-seat-candidate)  
`shrug_adr_elect` -> Election level with avg affidavit info. Each row is one election (one state-year-seat)  
`shrug_adr_cand` -> Candidate level with candidate affidavit info. Each row is one candidate (one state-year-seat-candidate)

## Basic Data Landscape

### Elections

We have access to state election level data for election results and candidate level affidavits. 

We first start by looking at the election level data. These include elections from 1979 to 2022.

```{r}
#| results: hold
print(nrow(shrug_triv_elect)) ## Total Elections in election data
print(nrow(shrug_triv_cand %>% select(sh_election_id) %>% unique())) ## Total Elections in candidate data
```

```{r}
#| results: hold
print(nrow(shrug_triv_elect %>% filter(bye_election == 0))) ## Total Elections in election data
print(nrow(shrug_triv_cand %>% filter(bye_election == 0) %>% select(sh_election_id) %>% unique())) ## Total Elections in candidate data
```

Therefore, we have a total of ~43,000 elections, which upon filtering to remove bye-poll elections leaves us with ~41,600 elections. 

We remove bye-polls because they usually have different, smaller sets of candidates and compaigns, and have no compulsion to have "None of the Above" as an option.


```{r}
election_is_nota <- shrug_triv_cand %>%
    filter(bye_election == 0) %>%
    group_by(sh_election_id, year) %>%
    summarise(is_nota = any(tolower(cand_name) == "nota" | tolower(cand_name) == "none of the above"), .groups = "drop")

election_is_nota %>%
    group_by(year) %>%
    summarise(
        perc_nota_elections = mean(is_nota),
        count_nota_elections = sum(is_nota),
        count_total_elections = n(),
    ) %>%
    print(n = 50)
```

Within these, as we can see, the elections with NOTA as an option start happening in 2013, and barring one bad data point, exist in every election since.

### Affidavits

<!-- 
```{r}
#| results: hold
print(nrow(shrug_adr_elect)) ## Total Elections in affidavit data
print(nrow(shrug_adr_cand %>% select(sh_election_id) %>% unique())) ## Total Elections in affidavit election data
```

```{r}
#| results: hold
print(nrow(shrug_adr_elect %>% filter(bye_election == 0))) ## Total Elections in affidavit data
print(nrow(shrug_adr_cand %>% filter(bye_election == 0) %>% select(sh_election_id) %>% unique())) ## Total Elections in affidavit election data
``` 
-->

Now, looking at the affidavit level data, we have information on affidavits from 2004 to 2017. While this limits the election data we are able to use, as we can see, it still gives us about 11,000 elections we can still analyse.

**Merging election level affidavit data**

```{r}
#| results: hold
#| message: false
adr_elect <- inner_join(shrug_adr_elect, shrug_triv_elect) %>% inner_join(election_is_nota)

adr_elect %>%
    group_by(year) %>%
    summarise(
        perc_nota_elections = mean(is_nota),
        count_nota_elections = sum(is_nota),
        count_total_elections = n(),
    ) %>%
    print(n = 50)
```

```{r}
adr_elect %>% group_by(is_nota) %>% summarise(n = n())
```

**Merging Candidate level affidavit data**

```{r}
#| message: false
adr_cand <- left_join(shrug_triv_cand, shrug_adr_cand, by = join_by("sh_cand_id", "sh_election_id"), suffix = c("", "_adr"), ) %>%
    inner_join(election_is_nota)

adr_cand %>%
    filter(!is.na(adr_cand_id)) %>%
    group_by(year) %>%
    summarise(
        mean_crimes = mean(num_crim, na.rm = T),
        median_crimes = median(num_crim, na.rm = T),
        max_crimes = max(num_crim, na.rm = T),
        perc_crime_major = mean(crime_major, na.rm = T),
        perc_crime_major_adr = mean(crime_major_adr, na.rm = T)
    ) %>%
    round(2)
```

## Estimations 

**Average NOTA share by state-year**

```{r}
#| message: false
adr_cand <- adr_cand %>%
    mutate(nota_cand = ifelse(tolower(cand_name) %in% c("nota", "none of the above"), 1, 0))

adr_cand %>%
    filter(nota_cand == 1) %>%
    group_by(eci_state_name, year) %>%
    summarise(
        median_vsp = median(vote_share_percentage),
        mean_vsp = mean(vote_share_percentage),
    ) %>%
    print(n=40)
```

At the state level, we can see that average percentage of votes polled as NOTA hoever around 1%, with interesting outliers in the case of the states of Bihar and Chhattisgarh - both poorer states with a (at least perceived) higher degree of corruption and maladministration.

**Constituency-year with NOTA votes higher than winning margin**

```{r}
#| message: false
election_nota_votes <- adr_cand %>%
    filter(nota_cand == 1) %>%
    mutate(
        nota_votes = votes,
        nota_vote_share = votes / valid_votes
    ) %>%
    select(sh_election_id, nota_votes, nota_vote_share)

nota_margin <- adr_cand %>%
    arrange(sh_election_id, desc(votes)) %>%
    group_by(sh_election_id) %>%
    mutate(
        win_margin = max(votes) - votes,
    ) %>%
    filter(win_margin != 0) %>%
    group_by(sh_election_id) %>%
    summarise(
        win_margin = min(win_margin)
    ) %>%
    inner_join(election_nota_votes) %>%
    mutate(diff_votes = win_margin - nota_votes) %>%
    arrange(diff_votes)

nota_margin %>% filter(diff_votes <= 0)
```

Surprisingly, of the 7,544 elections that have NOTA as an option, 474 or over 6% of the elections have a margin of victory that is less than the NOTA votes polled!

This points to the fact that some of these elections could actually be quite sensitive to the NOTA vote.

The NOTA vote represents something unique, it is someone taking the effort to the voting booth, stand in line for potentially hours, and then in some sense reveal a preference in a manner that is not really "relevant" to the election result. It is a costly signal with little to no benefit.

Let's now look at turnout trends, and vote-margins, pre and post NOTA

```{r}

```









```{r}
ggplot(data = nota_margin %>% filter(abs(diff_votes) < 500)) +
    geom_histogram(aes(x = diff_votes), binwidth = 5)
```


However, if elections are close, people seem to be voting for NOT nota so the election has more win margin(??)


## OTher tables 

Criminal affidavits per const


```{r}
election_crim <- adr_cand %>%
    filter(is_nota == 1) %>%
    group_by(sh_election_id) %>%
    summarise(
        sum_num_crim = sum(num_crim, na.rm = TRUE),
        num_crim = sum(num_crim > 0, na.rm = TRUE),
        median_num_crim = median(num_crim, na.rm = TRUE),
        mean_num_crim = mean(num_crim, na.rm = TRUE)
    ) %>%
    inner_join(election_nota_votes)


cm_nota <- election_crim %>%
    group_by(sum_num_crim) %>%
    summarise(
        median_nota_vote_share = median(nota_vote_share, na.rm = TRUE),
        mean_nota_vote_share = mean(nota_vote_share, na.rm = TRUE)
    )


ggplot(data = cm_nota) +
    geom_col(aes(x = sum_num_crim, y = median_nota_vote_share))
```


```{r}
major_crim <- adr_cand %>%
    filter(is_nota == 1) %>%
    group_by(sh_election_id) %>%
    summarise(
        crime_major = sum(crime_major_adr, na.rm = TRUE),
    ) %>%
    inner_join(election_nota_votes)


cm_nota <- major_crim %>%
    group_by(crime_major) %>%
    summarise(
        median_nota_vote_share = median(nota_vote_share, na.rm = TRUE),
        mean_nota_vote_share = mean(nota_vote_share, na.rm = TRUE)
    )


ggplot(data = cm_nota) +
    geom_col(aes(x = crime_major, y = median_nota_vote_share))
```


# Data

The data for this project comes from the following sources:

- PC and AC election results sourced from Lok Dhaba, TCPD, Ashoka University (@agarwal2021tcpd)
- Candidate affidavit data sourced from [myneta.info](myneta.info) by the Association for Democratic Reforms (ADR)


# References

::: {#refs .references}
:::