---
title: Final Project for Modeling Democracy
authors: Moksha Sharma, Divij Sinha
format: html
#bibliography: references.bib
execute:
  cache: true
echo: false
message: false
warning: false
---

# Introduction

Voting is costly: eligible voters need to collect information on candidates, formulate their preferences, and find the time to express these preferences at the ballot (Downs 1957, Feddersen and Sandroni 2006). Given these costs, political scientists and economists have offered many explanations for why citizens make the effort to vote despite their very low probability of casting the pivotal vote. Existing explanations including the policy benefits of voting for the voter (Downs 1957), psychological benefits from fulfilling one's civic duties (Riker and Ordeshook 1968), expressive benefits from supporting one's preferred candidates (Schuessler 2000), and signaling benefits from protest voting and abandoning preferred candidates to express discontentment (Kselman and Niou 2011).

This project focuses on the latter motivation for voting in India, and specifically for voting for no one. Many jurisdictions offer voters the option of formally expressing their discontentment with existing candidates by voting for none of them. This option to vote for “none of the above” (NOTA) was introduced in India in 2013 in an order of the Supreme Court. The stated goal of this policy was to boost turnout and convey to political parties which candidates they should not field in elections again (BBC 2013). In fact, the Supreme Court specifically argued that NOTA would give voters a proper mechanism for punishing candidates who are known to be associated with criminality (Goel 2018).

However, despite these lofty aims, NOTA carries no bite in its implementation in India (Association for Democratic Reforms 2024). While NOTA votes are counted and reported, they are essentially treated as invalid votes. The winning candidate is whomever gets the plurality of the remaining votes. While some countries require a bi-election to be conducted if the NOTA vote-share exceeds a certain threshold, there are no such requirements in India. Similarly, political parties are not required to ban candidates if, for instance, they win by a margin that is narrower than the share of NOTA votes. Thus, NOTA remains largely toothless and is little more than a costly signaling mechanism for casting protest votes.

Given this setup, this project seeks to look into the various ways NOTA has affected electoral outcomes in India. Descriptive work and existing research so far have generated various conflicting results. While Goel (2018) found no relationship between NOTA use and urban and literacy rates, increasing turnouts, or criminality in elections, Kailthya and Le (2025) found that NOTA was correlated with increased nighttime luminosity, which they considered to be a measure of democratic accountability. Using structural estimation, Ujhelyi, Chatterjee, and Szabo (2020) also argued that NOTA can improve political participation, albeit through a different mechanism, which is increasing turnout from voters who would have otherwise abstained.

To contribute to this growing literature, we use data from state legislature elections in India alongside data on criminal affidavits against political candidates. Using these data sources, we investigate the following questions regarding the usage of NOTA:

-   What is the average vote share received by NOTA in different states?

-   How has the margin of victory changed in elections since the introduction of NOTA?

-   How has the number of candidates in the elections changed since the introduction of NOTA?

-   How has NOTA affected the number of electoral candidates with criminal associations?

To briefly preview the results, we find that NOTA had modest to no effect on victory margins, turnout rates, and criminal associations of candidates.

The rest of this project proceeds as follows: the second section presents a basic decision model to show the circumstances under which a voter would be inclined to cast a NOTA vote, the third section gives a brief overview of the datasets we are using, and the fourth section shows our analysis and results. The final section concludes.

# Formal Model

We consider a simple decision model to show the conditions under which a voter would be inclined to cast a NOTA vote, even when they know that doing so will give the political party no formal incentive to change the candidate. In this simplified model, the Voter ($V$) can choose to either vote for their preferred candidate, choose NOTA, or abstain. The Voter's action set is thus $\{C,N,A\}$.

Let the policy benefit that the Voter gets from electing their preferred candidate be $\beta$. Formalising policy payoff in this way departs from the conventional technique for doing so, which represents it as the negative of the Euclidean distance between the voter's ideal point and the candidate's ideal point. The two techniques can be reconciled if we normalise the payoff from the opposing candidate as $0$ and think of $\beta>0$ as the policy gain from electing one's preferred candidate.

We now characterise the probabilities that $V$ assigns to their preferred candidate winning based on whether they vote for them or not. Let $\pi_v$ be the probability that the preferred candidate wins if $V$ votes for them and $\pi_0$ be the probability that they win if $V$ does not vote for them.

Finally, we characterise $c>0$ as the cost of voting net of the psychological gain from fulfilling one's civic duty, and $\alpha>0$ as the benefit of expressing discontentment to the candidates from voting for no one.

Thus, we get the following expected payoffs for the voter from their three possible strategies:

$$
\begin{aligned}
u_i(C) &= \pi_v\beta - c \\
u_i(N) &= \pi_0\beta +\alpha - c \\
u_i(C) &= \pi_0\beta \\
\end{aligned}
$$

For $V$ to choose NOTA, the following must hold:

$$
\begin{aligned}
\alpha &> c \\
\alpha &> (\pi_v -\pi_0)\beta
\end{aligned}
$$

From these conditions, we conclude the following:

-   As $(\pi_v-\pi_0) \rightarrow0$, the Voter feels more confident about their preferred candidate winning the election, and the threshold for casting NOTA lowers. In other words, if the Voter believes that the election is going to be won by their preferred candidate even without their vote, then they can afford to "waste" their vote by casting NOTA instead and enjoy expressive benefits instead. Thus, competitive constituencies might see more usage of NOTA.

-   As $\beta \rightarrow 0$, the Voter becomes more and more indifferent to the competing candidates or, alternatively, the candidates are largely undifferentiated to the Voter. Thus, the policy loss that they expect from not electing their preferred candidate is lower, which means that they are more inclined to vote NOTA.

# Data

The data for this project comes from the following sources:

-   PC and AC election results sourced from Lok Dhaba, TCPD, Ashoka University (@agarwal2021tcpd)
-   Candidate affidavit data sourced from [myneta.info](myneta.info) by the Association for Democratic Reforms (ADR)

```{r setup}
#| warning: false
#| output: false
library(tidyverse)
library(fixest)
library(gridExtra)
shrug_triv_elect <- read_csv("data/shrug-triv-elect-csv/trivedi_elections_clean.csv")
shrug_triv_cand <- read_csv("data/shrug-triv-cand-csv/trivedi_candidates_clean.csv")
shrug_adr_elect <- read_csv("data/shrug-adr-elections-ac-csv/affidavits_ac.csv")
shrug_adr_cand <- read_csv("data/shrug-adr-elections-cand-csv/affidavits_clean.csv")
```

`shrug_triv_elect` -\> Election level with winner info. Each row is one election (one state-year-seat)\
`shrug_triv_cand` -\> Candidate level with candidate info. Each row is one election candidate (one state-year-seat-candidate)\
`shrug_adr_elect` -\> Election level with avg affidavit info. Each row is one election (one state-year-seat)\
`shrug_adr_cand` -\> Candidate level with candidate affidavit info. Each row is one candidate (one state-year-seat-candidate)

## I. Elections Data

We have access to state election level data for election results and candidate level affidavits.

We first start by looking at the election level data. These include elections from 1979 to 2022. <!-- 
```{r}
#| results: hold
print(nrow(shrug_triv_elect)) ## Total Elections in election data
print(nrow(shrug_triv_cand %>% select(sh_election_id) %>% unique())) ## Total Elections in candidate data
```

```{r}
#| results: hold
print(nrow(shrug_triv_elect %>% filter(bye_election == 0))) ## Total Elections in election data
print(nrow(shrug_triv_cand %>% filter(bye_election == 0) %>% select(sh_election_id) %>% unique())) ## Total Elections in candidate data
``` 
-->

We start with a total of more than 43,000 elections. After filtering to remove bye-poll elections, we are left with approximately 41,600 elections[^1]. The plot below shows that after NOTA was introduced by the Supreme Court's order in 2013, every election has offered it to voters.

[^1]: We removed bye-polls because they the set of candidates varies and tends to be smaller, the electoral campaigns more limited, and no legal requirement for the voter to choose NOTA at the polls.

```{r}
election_is_nota <- shrug_triv_cand %>%
    filter(bye_election == 0) %>%
    group_by(sh_election_id, year) %>%
    summarise(is_nota = any(tolower(cand_name) == "nota" | tolower(cand_name) == "none of the above"), .groups = "drop")

election_is_nota %>%
    group_by(year) %>%
    summarise(
        perc_nota_elections = mean(is_nota),
        count_nota_elections = sum(is_nota),
        count_total_elections = n(),
    ) %>%
    ggplot(data = .) +
    geom_col(aes(x = year, y = count_total_elections, fill = "Total elections")) +
    geom_col(aes(x = year, y = count_nota_elections, fill ="Elections with NOTA")) + 
    theme_minimal()+
    labs(
        y = "Number of Elections",
        x = "Year",
        title = " Figure 1: Elections in the Dataset Before and After NOTA"
    )
```

## II. Merging with Criminal Affidavits Data

````{=html}
<!-- 
```{r}
#| results: hold
print(nrow(shrug_adr_elect)) ## Total Elections in affidavit data
print(nrow(shrug_adr_cand %>% select(sh_election_id) %>% unique())) ## Total Elections in affidavit election data
```

```{r}
#| results: hold
print(nrow(shrug_adr_elect %>% filter(bye_election == 0))) ## Total Elections in affidavit data
print(nrow(shrug_adr_cand %>% filter(bye_election == 0) %>% select(sh_election_id) %>% unique())) ## Total Elections in affidavit election data
``` 
-->
````

Now, looking at the affidavit level data, we have information on criminal affidavits from 2004 to 2017. This time-frame is shorter than the period over which we have electoral data, but we are still left with approximately 11,000 elections to analyse. Out of these elections, 6775 are pre-NOTA and 3346 are from after are it was introduced.

```{r}
#| results: hold
adr_elect <- inner_join(shrug_adr_elect, shrug_triv_elect) %>% inner_join(election_is_nota)

adr_elect %>%
    group_by(year) %>%
    summarise(
        perc_nota_elections = mean(is_nota),
        count_nota_elections = sum(is_nota),
        count_total_elections = n(),
    ) %>%
    ggplot(data = .) +
    geom_col(aes(x = year, y = count_total_elections, fill = "Total elections")) +
    geom_col(aes(x = year, y = count_nota_elections, fill = "Elections with NOTA")) +
    theme_minimal()+
    labs(
        y = "Number of Elections",
        x = "Year",
        title = "Figure 2: Elections Data Merged with Criminal Affidavits Data"
    )
```

```{r results='hide'}
adr_elect %>% group_by(is_nota) %>% summarise(n = n())
```

```{r}
adr_cand <- left_join(shrug_triv_cand, shrug_adr_cand, by = join_by("sh_cand_id", "sh_election_id"), suffix = c("", "_adr"), ) %>%
    inner_join(election_is_nota)
```

# Analysis

## I. Average NOTA Share by State-Year

For this section of the analysis, we use the cleaned dataset on elections (and not the much smaller version that was merged with the data on criminal affidavits). As we noted above, the criminal affidavits limit the temporal scope of the study. We return to the smaller dataset when we focus on criminality in a later section.

Figures 3 and 4 show the average share of NOTA ballots that were cast in each state over the course of the various elections in the dataset. We observe that on average, approximately 1% of votes were cast as NOTA. The two notable outliers are the states of Chhattisgarh and Bihar, which are both rural states with low literacy rates. The relatively higher frequency of NOTA use in these states has been covered by journalists, who in some cases have attributed it to "mistakes made by the voters due to low awareness" (Kumar 2014). In the case of Chhattisgarh, journalists have also written about Maoist groups encouraging voters to cast NOTA (Rai 2014).

```{r}
adr_cand <- adr_cand %>%
    mutate(nota_cand = ifelse(tolower(cand_name) %in% c("nota", "none of the above"), 1, 0)) 

adr_cand %>%
    filter(nota_cand == 1) %>%
    group_by(eci_state_name, year) %>%
    summarise(
        median_vsp = median(vote_share_percentage),
        mean_vsp = mean(vote_share_percentage),
    ) %>%
  invisible()

adr_cand %>%
    mutate(year = as.factor(year)) %>%
    filter(nota_cand == 1) %>%
    ggplot(data = .) +
    geom_density(aes(x = vote_share_percentage, fill = eci_state_name, alpha = 0.05)) +
    theme_minimal() +
    guides(alpha = "none")+
    labs(
        y = "Density",
        x = "NOTA Vote Share %",
        fill = "State",
        title = "Figure 3: Vote Share of NOTA by State"
    ) 

#plot1 <- last_plot()
```

```{r}
#| warning: false
election_nota_votes <- adr_cand %>%
    filter(nota_cand == 1) %>%
    mutate(
        nota_votes = votes,
        nota_vote_share = votes / valid_votes
    ) %>%
    select(sh_election_id, nota_votes, nota_vote_share, valid_votes)

#plot2 <- 
  ggplot(data = election_nota_votes, aes(x = nota_vote_share * 100)) +
    geom_density() +
    geom_vline(xintercept = 1, color = "red") +
    geom_vline(xintercept = 2, color = "red") +
    annotate("text", x = 1.1, y = 1, label = "1%", size = 3) +
    annotate("text", x = 2.1, y = 1, label = "2%", size = 3) +
    labs(
        x = "NOTA vote share (%)",
        y = "Density",
        title = "Figure 4: Distribution of NOTA vote share across elections"
    ) +
    theme_minimal()
```

```{r}
#grid.arrange(plot1, plot2, ncol = 2)
```

## **II. NOTA vs Margin of Victory**

Of the 7544 elections in the elections dataset that have NOTA as an option, 474 (or over 6%) have a margin of victory that is lower than the NOTA vote share. On the one hand, we could infer from this finding that the outcome of the election was sensitive to the NOTA vote share. On the other hand, based on Ujhelyi, Chatterjee, and Szabo (2020), this NOTA vote share may not be cutting into the vote share of the other candidates, and instead it may be representing the share of voters who would have otherwise abstained.

```{r}
win_margin <- adr_cand %>%
    arrange(sh_election_id, desc(votes)) %>%
    group_by(sh_election_id) %>%
    mutate(
        win_margin = max(votes) - votes,
        win_margin_perc = 100 * (win_margin / valid_votes)
    ) %>%
    filter(win_margin != 0) %>%
    group_by(sh_election_id) %>%
    summarise(
        win_margin = min(win_margin),
        win_margin_perc = min(win_margin_perc)
    )

nota_margin <- win_margin %>%
    inner_join(election_nota_votes) %>%
    mutate(diff_votes = win_margin - nota_votes) %>%
    arrange(diff_votes)

nota_margin %>% filter(diff_votes <= 0) %>% invisible()

ggplot(data = nota_margin %>% filter(diff_votes < 50000)) +
    geom_histogram(aes(x = diff_votes, fill = (diff_votes > 0)), binwidth = 500) +
  theme_minimal()+
  labs(x= "Difference between Margin of Victory and NOTA Vote Share",
       y="Number of Elections",
       title= "Figure 5: NOTA vs Margin of Victory")
```

## III. NOTA and Voter Turnout

To study the relationship between NOTA and voter turnout more rigorously, we use OLS. The standard errors are clustered at the state level.

In the first model specification, we regress turnout on the introduction of NOTA. We look at the decade before and after NOTA introduced and find that NOTA had no real impact on turnout trends. While there was a steep rise in the turnout rate until 2013, which when NOTA was introduced, we observe that the rate has since flattened.

Though this plateauing could be due to the introduction of NOTA, the more likely outcome is that both the introduction of NOTA and the rise in the turnout rate was driven by some unobserved confounder, such as the rise in political engagement following the 2013 anti-corruption protests in India. In other words, there is possibly an omitted variable bias here such that an unobserved confounder drove both electoral reform – like the introduction of NOTA – and increased voter participation. The subsequent flattening could then be explained as reversion to the mean.

In the second and third specifications, we regress win margins on NOTA and find that NOTA did not affect the average win margin. There is no such association even when we look at elections that were won by a narrow margin (1% or under).

Finally, in the fourth specification, we observe that NOTA did not have an effect on the number of candidates who were running in the elections either.

We have to keep in mind that the existence of NOTA is also a proxy for post-2013 in India, and this could also be an example of Duverger's law in practice, such that there is a general decrease in the number of candidates as campaigns are more professionally run, and there are more pre-poll alliances between parties, leading to a decrease in the number of candidates competing.

```{r}
m1_turnout <- shrug_triv_elect %>%
    filter(year >= 2004) %>%
    inner_join(election_is_nota) %>%
    mutate(year_2013 = year - 2013) %>%
    feols(fml = turnout_percentage ~ is_nota * year_2013 | ac_id, cluster = ~ pc01_state_id, data = .) #%>%
    #summary()
```

```{r}
m2_margin <- shrug_triv_elect %>%
    filter(year >= 2004) %>%
    inner_join(election_is_nota) %>%
    inner_join(win_margin) %>%
    mutate(year_2013 = year - 2013) %>%
    feols(fml = win_margin_perc ~ is_nota * year_2013 | ac_id, cluster = ~ pc01_state_id, data = .) #%>%
    #summary()
```

```{r}
m3_close_margin <- shrug_triv_elect %>%
    filter(year >= 2004) %>%
    inner_join(election_is_nota) %>%
    inner_join(win_margin) %>%
    filter(win_margin_perc <= 1) %>%
    mutate(year_2013 = year - 2013) %>%
    feols(fml = win_margin_perc ~ is_nota * year_2013 | ac_id, cluster = ~ pc01_state_id, data = .) #%>%
    #summary()
```

```{r}
m4_cand <- shrug_triv_elect %>%
    filter(year >= 2004) %>%
    inner_join(election_is_nota) %>%
    mutate(year_2013 = year - 2013) %>%
    feols(fml = n_cand ~ is_nota * year_2013 | ac_id, cluster = ~ pc01_state_id, data = .) #%>%
    #summary()
```

```{r}
library(knitr)

etable(
    m1_turnout, m2_margin, m3_close_margin, m4_cand,
    headers = c("Turnout %", "Win Margin %", "Win Margin (<=1%)", "Num Candidates"),
    fitstat = ~ n + r2,
    tex = FALSE # Forces fixest to output a clean data frame instead of latex/markdown
) %>% 
  kable()     # Knitr translates that data frame into a perfect Quarto HTML table
```

## IV. NOTA and Candidate Criminality

While the median candidate is not a criminal, we can see that the mean candidate is almost one, pulled up quite high by some "prolific" criminals.

The mean election candidate has \~0.2 crimes.

While NOTA has not had an impact on the number of candidates running in elections, there had been a generally modest increase in the number of criminal candidates every year, that seems to have stopped with NOTA. Moreover, there seems to also have been a slight drop in the number of criminal candidates at the advent of NOTA being introduced.

There also seems to have been a slight drop in NOTA vote share since the introduction of NOTA, in that when people were at their most politically engaged, they utilised NOTA more, but there seems to be no effect of criminality on NOTA vote share.

However, the wealth characteristics of a candidate (income, assets, liabilities, cash on hand) do not have correlation with how many NOTA votes an election gets.

```{r}
adr_cand %>%
    filter(!is.na(adr_cand_id)) %>%
    filter(nota_cand == 0) %>%
    group_by(year) %>%
    summarise(
        mean_crimes = mean(num_crim, na.rm = T),
        median_crimes = median(num_crim, na.rm = T),
        max_crimes = max(num_crim, na.rm = T),
        perc_crime_major = mean(crime_major, na.rm = T),
        perc_crime_major_adr = mean(crime_major_adr, na.rm = T)
    ) %>%
    round(2) %>% 
  invisible()
```

```{r}
adr_cand %>%
    filter(!is.na(adr_cand_id)) %>%
    filter(nota_cand == 0) %>%
    group_by(sh_election_id) %>%
    summarise(
        mean_crimes = mean(num_crim, na.rm = T),
        median_crimes = median(num_crim, na.rm = T),
        max_crimes = max(num_crim, na.rm = T),
        perc_crime_major = mean(crime_major, na.rm = T),
        perc_crime_major_adr = mean(crime_major_adr, na.rm = T)
    ) %>%
    summarise(
        median_mean_crimes = median(mean_crimes, na.rm = TRUE),
        median_median_crimes = median(median_crimes, na.rm = TRUE)
    ) %>% 
  invisible()
```

```{r}
m_crim_count <- adr_cand %>%
    filter(!is.na(adr_cand_id)) %>%
    filter(nota_cand == 0) %>%
    group_by(sh_election_id, pc01_state_id, ac_id) %>%
    summarise(
        num_criminals = sum(num_crim > 0, na.rm = TRUE),
        median_num_crim = median(num_crim, na.rm = TRUE),
        mean_num_crim = mean(num_crim, na.rm = TRUE)
    ) %>%
    ungroup() %>%
    inner_join(election_is_nota) %>%
    inner_join(shrug_triv_elect) %>%
    filter(year >= 2004) %>%
    mutate(
        year_2013 = year - 2013,
        crim_prop = 100 * (num_criminals / n_cand)
    ) %>%
    feols(fml = num_criminals ~ is_nota * year_2013 | pc01_state_id, cluster = ~pc01_state_id, data = .) 
```

```{r}
adr_election_crim <- adr_cand %>%
    filter(!is.na(adr_cand_id)) %>%
    group_by(sh_election_id, pc01_state_id, ac_id, year) %>%
    summarise(
        num_criminals = sum(num_crim > 0, na.rm = TRUE),
        median_num_crim = median(num_crim, na.rm = TRUE),
        mean_num_crim = mean(num_crim, na.rm = TRUE)
    ) %>%
    ungroup() %>%
    inner_join(election_nota_votes) %>%
    mutate(nota_vote_share = nota_vote_share * 100) %>%
    inner_join(shrug_triv_elect) %>%
    filter(year >= 2004) %>%
    mutate(
        year_2013 = year - 2013,
        crim_prop = 100 * (num_criminals / n_cand)
    )

m_crim_share <- feols(fml = nota_vote_share ~ crim_prop * year_2013 | pc01_state_id, cluster = ~pc01_state_id, data = adr_election_crim) 
```

```{r}
adr_election_assets <- adr_cand %>%
    filter(!is.na(adr_cand_id)) %>%
    group_by(sh_election_id, pc01_state_id, ac_id, year) %>%
    summarise(
        mean_liabilities = mean(liabilities, na.rm = T),
        mean_cash = mean(cash, na.rm = T),
        mean_income = mean(income, na.rm = T),
        mean_assets = mean(assets, na.rm = T),
        median_liabilities = median(liabilities, na.rm = T),
        median_cash = median(cash, na.rm = T),
        median_income = median(income, na.rm = T),
        median_assets = median(assets, na.rm = T),
    ) %>%
    ungroup() %>%
    inner_join(election_nota_votes) %>%
    mutate(nota_vote_share = nota_vote_share * 100) %>%
    filter(year >= 2004) %>%
    mutate(
        year_2013 = year - 2013,
        log_median_assets = log(median_assets + 1),
        log_median_liabilities = log(median_liabilities + 1),
        log_median_income = log(median_income + 1)
    )

m_assets<- feols(fml = nota_vote_share ~ log_median_assets * year_2013 | pc01_state_id, cluster = ~ pc01_state_id, data = adr_election_assets)
```

```{r}
m_liabilities <- feols(fml = nota_vote_share ~ log_median_liabilities * year_2013 | pc01_state_id, cluster = ~pc01_state_id, data = adr_election_assets) 
```

```{r}
etable(
    m_crim_count, m_crim_share, m_assets, m_liabilities,
    headers = c("Num Criminals", "NOTA Share (Crim)", "NOTA Share (Assets)", "NOTA Share (Liab)"),
    fitstat = ~ n + r2,
    tex = FALSE 
) %>% 
  kable()
```

# Conclusion

Overall, we find that NOTA had modest to no effect on victory margins, turnout rates, and criminal associations of candidates. These insignificant results are unsurprising, since the effort involved in voting NOTA is substantial. After all, even though 6,372,220 NOTA votes were cast in the last general election in 2024, they accounted for $0.99\%$ of the total vote share. Thus, while the decision to vote for NOTA is interesting political behaviour, it continues to be fringe behaviour nonetheless.

Another limitation of our study is that we did not account for many constituency level covariates that could have affected our results. That said, the effects we found were already insubstantial and insignificant, so adding further controls would not have made NOTA more significant.

Future research can look into the reasons why NOTA may be used with greater frequency in municipal and local elections, as voters may be able to better coordinate and send a stronger signal to the politicians and their parties. Such coordination may be harder to achieve for large-scale state-level or national elections.

----END----

NOTE: The rest is still WIP

However, if elections are close, people seem to be voting for NOT nota so the election has more win margin(??)

````{=html}
<!-- 
**Distribution of NOTA vote share**

```{r}
ggplot(data = nota_margin, aes(x = nota_vote_share * 100)) +
    geom_density(fill = "steelblue", alpha = 0.4, color = "steelblue4") +
    geom_vline(xintercept = 1, linetype = "dashed", color = "firebrick", linewidth = 0.7) +
    geom_vline(xintercept = 2, linetype = "dotted", color = "firebrick", linewidth = 0.7) +
    annotate("text", x = 1.1, y = Inf, label = "1%", vjust = 2, hjust = 0, color = "firebrick", size = 3) +
    annotate("text", x = 2.1, y = Inf, label = "2%", vjust = 2, hjust = 0, color = "firebrick", size = 3) +
    labs(
        x = "NOTA vote share (%)",
        y = "Density",
        title = "Distribution of NOTA vote share across elections"
    ) +
    theme_minimal()
```

The distribution is heavily right-skewed: the vast majority of elections see NOTA below 2%, but a non-trivial tail extends beyond that. The 1% and 2% reference lines put the "spoiler zone" in context — the elections flagged above as margin < NOTA votes live in this right tail.
-->
````

````{=html}
<!-- **NOTA vote share vs. winning margin**

```{r}
nota_margin_plot <- nota_margin %>%
    mutate(
        nota_vote_share_perc = nota_vote_share * 100,
        spoiler = diff_votes <= 0
    )

ggplot(data = nota_margin_plot, aes(x = nota_vote_share_perc, y = win_margin_perc, color = spoiler)) +
    geom_point(alpha = 0.3, size = 0.8) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black", linewidth = 0.8) +
    scale_color_manual(
        values = c("FALSE" = "steelblue", "TRUE" = "firebrick"),
        labels = c("FALSE" = "Margin > NOTA", "TRUE" = "Margin ≤ NOTA"),
        name = NULL
    ) +
    coord_cartesian(xlim = c(0, 8), ylim = c(0, 20)) +
    labs(
        x = "NOTA vote share (%)",
        y = "Win margin (% of valid votes)",
        title = "NOTA vote share vs. winning margin",
        subtitle = "Elections below the diagonal line have margin ≤ NOTA votes"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
```

Elections in red fall below the diagonal — their margin of victory was smaller than the NOTA vote. These are the elections where NOTA was structurally decisive, even though it mechanically could not affect the outcome. The clustering of most elections well above the diagonal confirms that this is a tail phenomenon, but the red points establish that it is not negligible.
 -->
````

## Other tables

Criminal affidavits per const

```{r}
adr_election_crim %>%
    group_by(num_criminals) %>%
    summarise(
        median_nota_vote_share = median(nota_vote_share, na.rm = TRUE),
        mean_nota_vote_share = mean(nota_vote_share, na.rm = TRUE)
    ) %>%
    ggplot(data = .) +
    geom_col(aes(x = num_criminals, y = median_nota_vote_share))
```

```{r}
major_crim <- adr_cand %>%
    filter(is_nota == 1) %>%
    group_by(sh_election_id) %>%
    summarise(
        crime_major = sum(crime_major_adr, na.rm = TRUE),
    ) %>%
    inner_join(election_nota_votes)


cm_nota <- major_crim %>%
    group_by(crime_major) %>%
    summarise(
        median_nota_vote_share = median(nota_vote_share, na.rm = TRUE),
        mean_nota_vote_share = mean(nota_vote_share, na.rm = TRUE)
    )


ggplot(data = cm_nota) +
    geom_col(aes(x = crime_major, y = median_nota_vote_share))
```

# References

Kumar, Sanjay (2014): "Is NOTA Serving Any Purpose?" Mint, 24 April, <http://www.livemint.> com/0pinion/b6ZPsuUYS2G3fWR2KItpaJ/Is NOTA-serving-any-purpose.html.

Rai, Chandan (2014): "Maoists Appear Divided Over Tactics During Elections 2014," India News Network, 18 April, <http://inn-live.blogs> p0t.c0.uk/2014/04/ma0ists-appear-divided over-tactics.html.
